Numerical simulation and entropy dissipative cure of the carbuncle instability for the shallow water circular hydraulic jump

### 2021-02-09 numerics_paper/version2/numerical_chj.tex
Write out full blended scheme.

\subsection{Roe's Riemann solver} \label{sec:roe}
The Roe Riemann solver is based on evaluating the flux Jacobian using Roe's average
\begin{align}\label{roe_average}
  \bar h_{i-1/2}=\frac{1}{2}(h_{i-1}+h_i), \qquad
  \hat u_{i-1/2}=\frac{\sqrt{h_{i-1}}u_{i-1}+\sqrt{h_i}u_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}, \qquad
  \hat v_{i-1/2}=\frac{\sqrt{h_{i-1}}v_{i-1}+\sqrt{h_i}v_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}.
\end{align}
The waves and speeds are given by the eigenvectors and eigenvalues of the approximate
flux Jacobian obtained using these averages, resulting in
\begin{align*}
  \hat\lambda_{i-1/2}^1=\hat u_{i-1/2}-g\sqrt{g\bar h_{i-1/2}}, \qquad
  \hat\lambda_{i-1/2}^2=\hat u_{i-1/2}, \qquad
  \hat\lambda_{i-1/2}^3=\hat u_{i-1/2}+g\sqrt{g\bar h_{i-1/2}}. \\
\end{align*}
and $\W_{i-1/2}^p=\alpha_{i-1/2}^p\hat r_{i-1/2}^p$, with
\begin{align*}
  \hat \bfr^1_{i-1/2} =
  \begin{bmatrix}
    1 \\
    \hat u_{i-1/2}-\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix},
  \qquad
  \hat \bfr^2_{i-1/2} =
  \begin{bmatrix}
    0\\
    0\\
    1
  \end{bmatrix},
  \qquad
  \hat \bfr^3_{i-1/2} =
  \begin{bmatrix}
    1 \\
    \hat u_{i-1/2}+\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix}
\end{align*}
and the factors $\alpha^p_\imh$ obtained by solving
\begin{align}\label{system_for_alphas}
  \left[\hat \bfr^1_{i-1/2} ~\hat \bfr^2_{i-1/2} ~\hat \bfr^3_{i-1/2}\right]
  \begin{bmatrix}
    \alpha^1_{i-1/2} \\
    \alpha^2_{i-1/2} \\
    \alpha^3_{i-1/2}
  \end{bmatrix}
  =\Delta Q_{i-1/2}:=q_i-q_{i-1}.
\end{align}

@@

In what follows, we use $\W^{p,\Rus}$ to denote the waves in the Rusanov
solver and simply $\W^p$ to denote the waves in the Roe solver.
%We close this section by remarking that the fluctuations associated with Rusanov's solver
%can be expressed in terms of waves and speeds from Roe's solver. To see this, first consider \eqref{rus_cons}
%and rewrite the right-going fluctuation as follows:
%\begin{align}\label{rus_rs_aux}
%  \A^{+,\Rus}\Delta Q_{i-1/2}:=\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\max}
%  = \bff(q_{i})-\bff(q_{i-1}) + \lambda_{i-1/2}^{\max}\left(\W_{i-1/2}^{1,\Rus}+\W_{i-1/2}^{2,\Rus}\right)
%  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus}.
%  %\A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
%\end{align}
%From \eqref{system_for_alphas} and \eqref{rusanov_waves} we get
%$\W^{1,\Rus}_{i-1/2}+\W^{2,\Rus}_{i-1/2}=q_i-q_{i-1}=\sum_p\W_{i-1/2}^p$.
%Using \eqref{rs_conservation}, \eqref{rus_rs_aux} becomes
%\begin{align*}
%  \lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\max}
%  = \sum_p \left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
%  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus},
%\end{align*}
%which implies that
%\begin{subequations}\label{fluct_rus}
%\begin{align}
%  \A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p,
%\end{align}
%and similarly,
%\begin{align}
%  \A^{-,\Rus}\Delta Q_{i+1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i+1/2}^p - \lambda_{i+1/2}^{\max}\right)\W_{i+1/2}^p.
%\end{align}
%\end{subequations}

@@

\subsection{Blended Riemann solver}\label{sec:blended_rs}
In this section, we use \eqref{Ri}, which is a normalized measurement of the entropy residual \eqref{ent_residual},
to blend Rusanov's and Roe's Riemann solvers. To do this, we define the blended numerical flux:
\begin{align}\label{num_flux_ev}
  F_{i+1/2} = \frac{\bff(\bfq_{i+1})+\bff(\bfq_i)}{2}
  - \frac{1}{2} \left( \theta_{i+1/2}\lambda_{i+1/2}^{\max} I + (1-\theta_{i+1/2})|\hat A_{i+1/2}| \right)(\bfq_{i+1}-\bfq_{i}).
\end{align}
Here $I\in\mathbb{R}^{3\times 3}$ is the identity matrix,
$\theta_{i+1/2}=\max\{\theta_{i+1},\theta_i\}$, $\lambda_{i+1/2}^{\max}$ is the
upper bound on the wave speed used in Rusanov's Riemann solver (see \S \ref{sec:rusanov}),
and $\hat A_{i+1/2}$ is Roe's averaged flux Jacobian (see \S \ref{sec:roe}).

The first-order method is given simply by \eqref{first-order_FV}
with the numerical flux $F_{i+1/2}$ given by \eqref{num_flux_ev}.
In smooth regions $\theta_{i+1/2}\approx 0$ so the numerical flux is essentially Roe's flux.
Near discontinuities or steep gradients, $\theta_{i+1/2}$ becomes large. In the limit $\theta_{i+1/2}\approx 1$
the numerical flux is essentially Rusanov's flux.

The first-order method can also be written in terms of fluctuations via \eqref{first-order_via_fluct}.
To do this we consider the waves ($\W^p_{i\pm 1/2}$, with $p=1,2,3$)
from Roe's solver and define
\begin{align}\label{lambda_p}
  \lambda_{i+1/2}^p := \theta_{i+1/2}\lambda_{i+1/2}^{\max} + (1-\theta_{i+1/2})|\hat \lambda_{i+1/2}^p|,
\end{align}
where $\hat\lambda_{i+1/2}^p$ is the $p$-th eigenvalue of Roe's solver corresponding to the
interface $i+1/2$. Note that $\lambda_{i+1/2}^p\geq 0$.
The fluctuations are then given by
\begin{subequations}\label{ev_fluctuations}
\begin{align}
  \apdq_\imh & = \frac{1}{2}\sum_p \left(\hat\lambda_{i-1/2}^p + \lambda_{i-1/2}^p\right)\W_{i-1/2}^p, \\
  \amdq_\iph & = \frac{1}{2}\sum_p \left(\hat\lambda_{i+1/2}^p - \lambda_{i+1/2}^p\right)\W_{i+1/2}^p.
\end{align}
\end{subequations}

To implement a second-order LWL scheme based on this solver, in addition to
these fluctuations we must define a set of waves and corresponding speeds.
Using only the waves from the Roe solver, it is in general not possible
to choose speeds that yield the fluctuations \eqref{ev_fluctuations}.
%In the framework of LeVeque, for a given interface $i+1/2$,
%we would need to consider a wave speed $s_{i+1/2}^p$ so that
%\begin{align*}
%  \hat \lambda_{i+1/2}^p+\lambda_{i+1/2}^p = s_{i+1/2}^p+|s_{i+1/2}^p|, \qquad
%  \hat \lambda_{i+1/2}^p-\lambda_{i+1/2}^p = s_{i+1/2}^p-|s_{i+1/2}^p|,
%\end{align*}
%which is not possible to obtain in general (since the problem is overdetermined).
Instead, we consider
$s_{i+1/2}^p=\sgn(\hat\lambda_{i+1/2}^p)\lambda_{i+1/2}^p$
 and study the consequences of this choice once the second-order corrections are applied.
Here $\sgn(z)$ is the sign function of $z$.

The second-order LWL method for the one-dimensional problem is given by
\eqref{second-order_via_fluct} and \eqref{correction-fluxes}.
To understand the effect of choosing $s_{i+1/2}^p=\sgn(\hat\lambda_{i+1/2}^p)\lambda_{i+1/2}^p$
as the speed of $\W_{i+1/2}^p$,
let us assume that no limiter is applied; i.e., $\tilde\W_{i\pm 1/2}^p\equiv \W_{i\pm 1/2}^p$.
Then \eqref{second-order_via_fluct} can be written as
\begin{align*}
  q_i^{n+1}=q_i^n
  &-\frac{\Delta t}{2 \Delta x}
  \sum_p
  \left[\left(\hat\lambda_{i+1/2}^p -\frac{\Delta t}{\Delta x}(\lambda_{i+1/2}^p)^2 \right)\W_{i+1/2}^p
  +
  \left(\hat\lambda_{i-1/2}^p +\frac{\Delta t}{\Delta x}(\lambda_{i-1/2}^p)^2 \right)\W_{i-1/2}^p\right],
\end{align*}
which is just the standard Lax-Wendroff method written wave propagation form.
Now taking $\theta_{i\pm 1/2}=0$ corresponds to using a Roe solver while taking
$\theta_{i\pm 1/2}=1$ corresponds to using a Rusanov solver.

Note that if $\theta_{i\pm 1/2}=0$ we recover the standard second-order Lax-Wendroff method based on
Roe's Riemann solver.
In the numerical experiments in \S\ref{sec:num}, we obtain second-order experimental
order of convergence (EOC) for problems with smooth solution, which is expected since
$\theta_{i\pm 1/2}\approx 0$ in smooth regions.
Similarly, if $\theta_{i\pm 1/2}=1$, we recover the Lax-Wendroff method based on Rusanov's Riemann solver,
which can be seen by writing the fluctuations in Rusanov's method in terms of Roe's waves and speeds.

>>>>

\subsection{Roe's Riemann solver} \label{sec:roe}
The Roe Riemann solver is based on evaluating the flux Jacobian using Roe's average
\begin{align}\label{roe_average}
  \bar h_{i-1/2}=\frac{1}{2}(h_{i-1}+h_i), \qquad
  \hat u_{i-1/2}=\frac{\sqrt{h_{i-1}}u_{i-1}+\sqrt{h_i}u_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}, \qquad
  \hat v_{i-1/2}=\frac{\sqrt{h_{i-1}}v_{i-1}+\sqrt{h_i}v_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}.
\end{align}
The waves and speeds are given by the eigenvectors and eigenvalues of the approximate
flux Jacobian obtained using these averages, resulting in
\begin{align*}
  \hat\lambda_{i-1/2}^1=\hat u_{i-1/2}-g\sqrt{g\bar h_{i-1/2}}, \qquad
  \hat\lambda_{i-1/2}^2=\hat u_{i-1/2}, \qquad
  \hat\lambda_{i-1/2}^3=\hat u_{i-1/2}+g\sqrt{g\bar h_{i-1/2}}. \\
\end{align*}
and $\W_{i-1/2}^{p,\Roe}=\alpha_{i-1/2}^p\hat r_{i-1/2}^p$, with
\begin{align*}
  \hat \bfr^1_{i-1/2} =
  \begin{bmatrix}
    1 \\
    \hat u_{i-1/2}-\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix},
  \qquad
  \hat \bfr^2_{i-1/2} =
  \begin{bmatrix}
    0\\
    0\\
    1
  \end{bmatrix},
  \qquad
  \hat \bfr^3_{i-1/2} =
  \begin{bmatrix}
    1 \\
    \hat u_{i-1/2}+\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix}
\end{align*}
and the factors $\alpha^p_\imh$ obtained by solving
\begin{align}\label{system_for_alphas}
  \left[\hat \bfr^1_{i-1/2} ~\hat \bfr^2_{i-1/2} ~\hat \bfr^3_{i-1/2}\right]
  \begin{bmatrix}
    \alpha^1_{i-1/2} \\
    \alpha^2_{i-1/2} \\
    \alpha^3_{i-1/2}
  \end{bmatrix}
  =\Delta Q_{i-1/2}:=q_i-q_{i-1}.
\end{align}

@@

%We close this section by remarking that the fluctuations associated with Rusanov's solver
%can be expressed in terms of waves and speeds from Roe's solver. To see this, first consider \eqref{rus_cons}
%and rewrite the right-going fluctuation as follows:
%\begin{align}\label{rus_rs_aux}
%  \A^{+,\Rus}\Delta Q_{i-1/2}:=\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\max}
%  = \bff(q_{i})-\bff(q_{i-1}) + \lambda_{i-1/2}^{\max}\left(\W_{i-1/2}^{1,\Rus}+\W_{i-1/2}^{2,\Rus}\right)
%  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus}.
%  %\A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
%\end{align}
%From \eqref{system_for_alphas} and \eqref{rusanov_waves} we get
%$\W^{1,\Rus}_{i-1/2}+\W^{2,\Rus}_{i-1/2}=q_i-q_{i-1}=\sum_p\W_{i-1/2}^p$.
%Using \eqref{rs_conservation}, \eqref{rus_rs_aux} becomes
%\begin{align*}
%  \lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\max}
%  = \sum_p \left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
%  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus},
%\end{align*}
%which implies that
%\begin{subequations}\label{fluct_rus}
%\begin{align}
%  \A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p,
%\end{align}
%and similarly,
%\begin{align}
%  \A^{-,\Rus}\Delta Q_{i+1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i+1/2}^p - \lambda_{i+1/2}^{\max}\right)\W_{i+1/2}^p.
%\end{align}
%\end{subequations}

@@

\subsection{Blended Riemann solver}\label{sec:blended_rs}
In what follows, we use $\W^{p,\Rus}$ to denote the waves in the Rusanov
solver and $\W^{p,\Roe}$ to denote the waves in the Roe solver.  We propose
a new scheme consisting of the second-order LWL method \eqref{second-order_via_fluct} using
the first-order fluxes
\begin{align}
  \bfF_{i+1/2} = \frac{\bff(\bfq_{i+1})+\bff(\bfq_i)}{2}
  - \frac{1}{2} \left( \theta_{i+1/2}\lambda_{i+1/2}^{\max}\mathbb{I} + (1-\theta_{i+1/2})|\hat A_{i+1/2}| +\lambda_{i+1/2}^{\min}\mathbb{I}\right)(\bfq_{i+1}-\bfq_{i})
\end{align}
and correction fluxes \eqref{correction-fluxes} where
\begin{align} \label{s_p}
    s_{i+1/2}^p=\sgn(\hat\lambda_{i+1/2}^p)\lambda_{i+1/2}^p
\end{align}
with
\begin{align}\label{lambda_p}
  \lambda_{i+1/2}^p := \theta_{i+1/2}\lambda_{i+1/2}^{\max} + (1-\theta_{i+1/2})|\hat \lambda_{i+1/2}^p| + \lambda_\iph^{\min}.
\end{align}
The term $\lambda_\iph^{\min}$ is used to ensure that the scheme would be entropy stable
if the second-order corrections were neglected.  The normalized entropy
residual $\theta$ controls whether the scheme behaves more like that of Rusanov
or Roe.  Specifically, if $\lambda_\iph^{\min}=0$ and $\theta_\iph=0$, this scheme
is equivalent to that of Rusanov, while if $\lambda_\iph^{\min}=0$ and $\theta_\iph=1$, 
this scheme is equivalent to that of Roe.

In this section, we use \eqref{Ri}, which is a normalized measurement of the entropy residual \eqref{ent_residual},
to blend Rusanov's and Roe's Riemann solvers. To do this, we define the blended numerical flux:
\begin{align}\label{num_flux_ev}
  F_{i+1/2} = \frac{\bff(\bfq_{i+1})+\bff(\bfq_i)}{2}
  - \frac{1}{2} \left( \theta_{i+1/2}\lambda_{i+1/2}^{\max} I + (1-\theta_{i+1/2})|\hat A_{i+1/2}| \right)(\bfq_{i+1}-\bfq_{i}).
\end{align}
Here $I\in\mathbb{R}^{3\times 3}$ is the identity matrix,
$\theta_{i+1/2}=\max\{\theta_{i+1},\theta_i\}$, $\lambda_{i+1/2}^{\max}$ is the
upper bound on the wave speed used in Rusanov's Riemann solver (see \S \ref{sec:rusanov}),
and $\hat A_{i+1/2}$ is Roe's averaged flux Jacobian (see \S \ref{sec:roe}).
This flux is equal to the Rusanov flux if $\theta_\iph=0$ and the Roe flux if $\theta_\iph=1$.
We expect that $\theta_\iph\approx 0 $ in smooth regions, while $\theta_\iph\approx 1$ near
shocks.

%The first-order method is given simply by \eqref{first-order_FV}
%with the numerical flux $F_{i+1/2}$ given by \eqref{num_flux_ev}.

The first-order method can also be written in terms of fluctuations via \eqref{first-order_via_fluct}.
To do this we define the blended wave speed
\begin{align}\label{lambda_p}
  \lambda_{i+1/2}^p := \theta_{i+1/2}\lambda_{i+1/2}^{\max} + (1-\theta_{i+1/2})|\hat \lambda_{i+1/2}^p|,
\end{align}
where $\hat\lambda_{i+1/2}^p$ is the $p$-th eigenvalue of Roe's solver corresponding to the
interface $i+1/2$. Note that $\lambda_{i+1/2}^p\geq 0$.
The fluctuations are then given by
\begin{subequations}\label{ev_fluctuations}
\begin{align}
  \apdq_\imh & = \frac{1}{2}\sum_p \left(\hat\lambda_{i-1/2}^p + \lambda_{i-1/2}^p\right)\W_{i-1/2}^{p,\Roe}, \\
  \amdq_\iph & = \frac{1}{2}\sum_p \left(\hat\lambda_{i+1/2}^p - \lambda_{i+1/2}^p\right)\W_{i+1/2}^{p,\Roe}.
\end{align}
\end{subequations}
These can be used in \eqref{first-order_via_fluct} to obtain a first-order scheme.
To implement the correction fluxes required for the second-order scheme \eqref{second-order_via_fluct},
we must also define a set of waves and corresponding speeds.
Using only the waves from the Roe solver, it is in general not possible
to choose speeds that yield the fluctuations \eqref{ev_fluctuations}.
%In the framework of LeVeque, for a given interface $i+1/2$,
%we would need to consider a wave speed $s_{i+1/2}^p$ so that
%\begin{align*}
%  \hat \lambda_{i+1/2}^p+\lambda_{i+1/2}^p = s_{i+1/2}^p+|s_{i+1/2}^p|, \qquad
%  \hat \lambda_{i+1/2}^p-\lambda_{i+1/2}^p = s_{i+1/2}^p-|s_{i+1/2}^p|,
%\end{align*}
%which is not possible to obtain in general (since the problem is overdetermined).
Instead, we set
\begin{align} \label{s_p}
    s_{i+1/2}^p=\sgn(\hat\lambda_{i+1/2}^p)\lambda_{i+1/2}^p
\end{align}
where $\sgn(z)$ is the sign function of $z$.
Using this in \eqref{correction-fluxes} and \eqref{ev_fluctuations} in 
\eqref{second-order_via_fluct}, we obtain (in the absence of limiting)
the second-order scheme
\begin{align*}
  q_i^{n+1}=q_i^n
  &-\frac{\Delta t}{2 \Delta x}
  \sum_p
  \left[\left(\hat\lambda_{i+1/2}^p -\frac{\Delta t}{\Delta x}(\lambda_{i+1/2}^p)^2 \right)\W_{i+1/2}^p
  +
  \left(\hat\lambda_{i-1/2}^p +\frac{\Delta t}{\Delta x}(\lambda_{i-1/2}^p)^2 \right)\W_{i-1/2}^p\right],
\end{align*}
Note that if $\theta_{i\pm 1/2}=0$ we recover the standard second-order Lax-Wendroff method based on
Roe's Riemann solver.
Similarly, if $\theta_{i\pm 1/2}=1$, we recover the Lax-Wendroff method based on Rusanov's Riemann solver,
which can be seen by writing the fluctuations in Rusanov's method in terms of Roe's waves and speeds.
